Tree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest

### 2020-07-23 manuscript/climate_sensitivity_draft.tex
more rooting depth refs

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Tree size and hydraulic traits shape drought responses in a temperate broadleaf forest},
            pdfauthor={Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{apalike}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Tree size and hydraulic traits shape drought responses in a temperate
broadleaf forest}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine
Herrmann, Joseph Zailaa, Atticus Stovall, Neil Pederson, Lawren Sack,
Krista Anderson-Teixeira}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
    \date{}
    \predate{}\postdate{}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\usepackage{float}
\usepackage{booktabs}
\usepackage{pdflscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\usepackage{caption}
\captionsetup[table]{font=small}
\captionsetup[figure]{font=small}
\captionsetup[table]{labelformat=empty}
\usepackage{dcolumn}

\begin{document}
\maketitle

\subsubsection{Summary}\label{summary}

Predicting forest responses to drought is an increasingly critical task
under climate change effects. Part of the problem is due to the lack of
studies analyzing the confluence of leaf hydraulic traits with
biophysical parameters. In this study, we analyze the interaction
between these two trait groups using forest census data from a 25.6-ha
ForestGEO plot in Virginia (USA). Drought periods were defined by both
Palmer Drought Severity Indices (PDSI) and their identification from
tree-ring records for 12 species representing 97\% of woody
productivity. Each drought scenario (1964-66, 1977, 1999), along with
the overall trend, was then tested against leaf hyraulic trait
measurements and microhabitat biophysical data. Individual-level growth
responses to the three individual droughts were stronger among taller
trees in dominant canopy positions, those in wetter microsites, and for
more drought-sensitive species as assessed by leaf traits (turgor loss
at less negative leaf water potential, greater shrinkage with leaf
dehydration), with substantial variation in the best predictor variables
across given droughts. We conclude that when droughts occur, large
dominant trees, drought-sensitive species, and individuals in wetter
microhabitats are likely to be most strongly affected.

\emph{The Summary for research papers, which must be usable as a stand-
alone document, must not exceed 200 words and should be organized using
four bullet points to indicate: (1) the research conducted, including
the rationale, (2) methods, (3) key results, and (4) the main
conclusion, including the key points of discussion. It should not
contain citations of other papers.}

\subsubsection{Introduction}\label{introduction}

Understanding how and why trees respond to drought is critical to
predicting forest drought responses and climate change feedbacks. (1
paragraph on this- I have lots of content on this and will add later.)

{[}Understanding forest responses to drought requires increased
functional understanding of the factors that confer individual-level
vulnerability or resistance.{]} Forests are diverse in terms of tree
sizes and functional traits, and it is known that trees varying in size
and functional traits respond differently to drought (e.g.,
\citep{bennett_larger_2015}; REFS). Therefore, in order to understand
whole-forest response to drought, we need a functional understanding of
how responses vary by tree size, microhabitat, and species. There are 3
fundamental questions that must be addressed:

\emph{First, what drives the observed tendency for large trees to suffer
more during drought?}\\
\citet{bennett_larger_2015} showed that in forests globally, large trees
suffer greater growth reductions during drought. However, this analysis
quantified tree size based on DBH, which has no direct mechanistic
meaning. This study proposed two major mechanisms (besides insects) for
the observed greater drougth growth reductions of large trees: (1)
inherently greater biophysical challenge of being tall; (2) greater
exposure of the crowns of large trees. Counteracting these effects, (3)
the larger root systems of larger trees may confer an advantage in terms
of allowing greater access to water, but it appears that this effect is
usually insufficient to offset the costs of height and/or crown
exposure.

Canopy trees have lower drought resistance because they are exposed to
higher solar radiation, greater wind speeds, and lower humidity.
Alternatively, the generally supressed status of subcanopy trees may be
insufficient to override the benefits of their buffered environment
during drought.

\emph{Second, how do species' traits - alone and in interaction with
tree size - influence drought response?} Analyzing drought responses on
the species level does not fully explain mechanisms and is not feasible
in diverse forests. The solution is a trait-based approach. Leaf
hydraulic traits hold more promise than more
commonly/traditionally-measured traits such as wood density and specific
leaf area (SLA)
(\href{https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229}{Medeiros
et al. 2019}).

Commonly measured traits including wood density (REFS), leaf mass per
area \citep{abrams_adaptations_1990} \citep{guerfel_impacts_2009}, and
ring porosity (Elliot et al. 2015, Friedrichs et al. 2009) have been
linked to drought responses and likely correlated with drought
resistance in this forest. However, we hypothesize that leaf hydraulic
traits such as leaf area shrinkage upon dessication (PLA) and turgor
loss point (TLP), which are emerging as potentially more informative
traits but whose effect on drought resistance has never been tested
(CONFIRM), will prove better predictors.

Moreover, there may be an interaction between traits and tree size. It
is possible that the pattern observed by \citet{bennett_larger_2015}
could be caused by smaller trees being more drought resistant.
Alternatively, larger trees may have more drought-resistant traits as
adpatations to greater biophysical challenges.

\emph{Third, are responses similar or variable across individual drought
years?} Droughts are rarely explicitly defined in ecological studies
\citep{slette_how_2019}, yet no two droughts are the same.

We need to understand the the factors confirming drought vulnerability
or resistance not only for extreme droughts with dramatic impacts on
tree growth and mortality, which tend to dominate the literature
\citep{bennett_larger_2015} (REFS), but also for more modest but
frequent droughts--e.g., those with historical return intervals on the
order of a decade.

Here, we combine tree-ring records covering three droughts (1964-66,
1977, 1999), species functional and hydraulic trait measurments, and
forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to
test a series of hypotheses and associated specific predictions (Table
1) designed to yield functional understanding of how tree size,
microenrivonment, and species' traits collectively shape drought
responses. First, we focus on the role of tree size and its interaction
with microenvironement. We confirm that, consistent with most forests
globally \citep{bennett_larger_2015}, larger-diameter trees have lower
drought resistance in this forest, which is in an ecoregion represented
by only one study in Bennett et al (2015) (H1.0). We then test
hypotheses designed to disentangle the relative importance of tree
height (H1.1), crown exposure (H1.2), and root water access, which
should be greater for larger trees in dry but not perpetually wet
microsites (H1.3). Second, we focus on the role of species' functional
and hydraulic traits and their interaction with tree height. We
hypothesize that drought resistance will follow predicted and observed
patterns in relation to wood density, specific leaf area, and ring
porosity (P2.1a-c), but that leaf hydraulic traits such as leaf area
shrinkage upon dehydration and turgor loss point will prove better
predictors (P2.1d-e). We then test whether these traits correlate with
tree height (P2.2), potentially driving the observed tendency for taller
trees to suffer more during drought (P2.3). Finally, we focused on
variability among droughts, asking how community resistance varied
across droughts (H3.1) and whether the factors confirming vulnerability
or resistance varied across droughts (H3.2).

\emph{Let's put Table 1 here. We need to give the hypotheses/
predictions here. It seems counterintuitive to put the results right up
front like that, but it actually tends to be a higher-impact style of
writing, and recommended
\href{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005619}{here}.}

\textbf{Table 1. Summary of hypotheses, predictions and results}
\includegraphics[width=5.20833in]{tables_figures/Table1.png}

We count predictions as fully supported (or rejected) when the direction
of response matches (or contradicts) the prediction. H2.3-H3.2 are based
on if models containing species traits in H2.1 had dAIC\textgreater{}=2
relative to the appropriate null model or to any alternative
multivariate model within 2 dAIC. Parentheses indicate that predictions
were partially supported (or rejected). In other words, parentheses
indicate when the direction of response matched (or contradicted) the
prediction in some but not all models with dAIC\textless{}2 relative to
the appropriate null model. With categorical variables such as crown
position, a ``(yes)'' notation describes when the trend matched the
prediction, but it wasn't significant A ``(yes/no) or (''no/yes``)''
indicates tendencies in opposite directions in univariate tests versus
the best full models, respectively. ``(yes)

\subsubsection{Materials and Methods}\label{materials-and-methods}

\emph{Study site} Research was conducted at the 25.6 ha ForestGEO
(Global Earth Observatory) study plot at the Smithsonian Conservation
Biology Institute (SCBI) in Virginia, USA (38°53'36.6``N, 78°
08'43.4''W) \citep{andersonteixeira_ctfs-forestgeo:_2015}. SCBI is
located in the central Appalachian Mountains at the northern edge of
Shenandoah National Park. Elevations range from 273-338m above sea level
\citep{gonzalezakre_patterns_2016} with a topographic relief of 65m
\citep{bourg_initial_2013}. Dominant species include \emph{Liriodendron
tulipifera}, oaks (\emph{Quercus} spp.), and hickories (\emph{Carya}
spp.).

\emph{Data collection and preparation} The SCBI ForestGEO plot was
censused in 2008, 2013, and 2018 following standard ForestGEO protocols,
whereby all free-standing woody stems \textgreater{}= 1cm diameter at
breast height (DBH) were mapped, tagged, measured at DBH, and identified
to species \citep{condit_tropical_1998}. From this census data, we used
measurements of DBH from 2008 to calculate historical DBH (described
below), along with tree location in the plot to determine the
topographic wetness index. Furthermore, we used data for all stems
\textgreater{}= 10cm to analyze functional trait composition relative to
tree height. \emph{I added the last sentence-- not sure if you've
described that below yet. I don't think you need to note here which data
we use, so long as that's described in the portions where you describe
the analysis.}

We analyzed tree-ring data from the twelve species contributing most to
woody aboveground net primary productivity (ANPP), which together
comprised 97\% of study plot ANPP between 2008 and 2013
\citep{helcoski_growing_2019}. Cores were obtained in 2010-2011 or
2016-2017 from a breast height of 1.3m using a 5mm increment borer. In
2010-2011, cores were collected from randomly selected live trees of
species with at least 30 individuals of DBH \textgreater{}= 10cm
\citep{bourg_initial_2013}. In 2016-2017, cores were collected from all
trees found dead in the annual mortality census
\citep{gonzalezakre_patterns_2016}. Cores were sanded, measured, and
cross-dated using standard procedures, as detailed in
\citep{helcoski_growing_2019}. The resulting chronologies have been
published in association with \citet{helcoski_growing_2019}: (ITRDB;
GitHub/Zenodo). \emph{is this a database? Yes. Ryan submitted the data
but I don't think its posted yet. We should also cite GitHub/Zenodo
here. I'll come back to that. }

Height measurements (n=\# trees) were taken by several researchers
between 2012 to 2019, and are archived in a public
\href{https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights}{GitHub
repository}. Measurement methods included manual
\citep[NEON]{stovall_assessing_2018}, digital rangefinders
\citep{andersonteixeira_size-related_2015}, and automatic LiDAR
\citep{stovall_terrestrial_2018}. Rangefinders either used the tangent
method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon
ForestryPro) for calculating heights. The associated errors for using
either method were acknowledged \citep{larjavaara_measuring_2013}.
Species-specific height allometries were developed (Table S\# -
\textbf{ADD THIS TABLE TO SI}). For species that didn't have enough
height measurements, heights were calculated from equations derived from
all species in the study.

For each tree, we combined tree-ring records and allometric equations of
height and bark thickness to retroactively calculate DBH and estimate
height for the years 1950-2009. Prior DBH was estimated using the
following equation, using 2008 as the earliest year for having reliable
DBH measurements:

\[ diamYEAR = dbh2008 - 2*(bark.depth2008) - 2*\Sigma(ring.widthYEAR:ring.width2008) + 2*(bark.depthYEAR) \]

Here, \emph{ring.width} was measured from cores. Bark thickness was
estimated from species-specific allometries based on the bark thickness
data of \citep{andersonteixeira_size-related_2015}. Specifically, we
used linear regression equations on log-transformed data to relate bark
thickness to DBH (Table S\#- \textbf{create table to give these
equations in SI}) and then used these to estimate bark thickness based
on DBH.

Crown positions were recorded in the field during the growing season of
2018 following the crown position protocol from
\citep{jennings_assessing_1999}, whereby positions were ranked as
dominant, codominant, intermediate, or suppressed. As there was no way
to retroactively estimate crown position, we assumed that 2018 crown
position was reflective of each tree's position over the past 60 years.
While some trees undoubtedly changed position, an analysis of crown
position relative to height (Fig. XX) and height change since
\emph{1959} indicated that change was likely slow. {[}\textbf{work on
this-- provide details?}{]}

Topographic wetness index (TWI) was calculated using the
\citep{R-dynatopmodel} package in R.

\begin{figure}[htbp]
\centering
\includegraphics[width=5.20833in]{tables_figures/Figure3.png}
\caption{Location of cored trees}
\end{figure}

Hydraulic traits were collected from SCBI and are summarized in Table 1.
In August 2018, we collected leaf samples from three individuals of each
species \ldots{} (\textbf{Nobby's description of methods for the
following (see word document}) 1. PLA 2. LMA 4. Wood density 5. TLP

\textbf{Table 2. Species analyzed here, listed in descending order of
ANPP\_stem. n cores and DBH range represented, and species traits}
{[}*This replaces/combines the two remaining tables in this section.
Suggested columns, with those to include only if they fit in
parentheses: species, (stems \textgreater{}=10 cm per ha in plot),
(ANPP\_stem), n cores, DBH range of cores, (n cores in each crown
position) species means for each trait{]}

\begin{verbatim}
## Warning: package 'kableExtra' was built under R version 3.4.4
\end{verbatim}

\textbackslash{}begin\{table\}{[}!h{]}

\textbackslash{}caption\{\label{tab:Table2}Overview of analyzed species,
detailing DBH mean and range of cored trees, the number of cores
represented by each crown position of each species, and mean hydraulic
trait measurements. Units of measurements are in mm (DBH), \% (PLA),
g/m2 (LMA), MPa (TLP), and g/cm3 (WD).\} \centering

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{setspace}\onehalfspacing
\usepackage{lineno}
\linenumbers
\usepackage{float}
\usepackage{booktabs}
\usepackage{pdflscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\usepackage{caption}
\captionsetup[table]{font=small}
\captionsetup[figure]{font=small}
\captionsetup[table]{labelformat=empty}
\captionsetup[figure]{labelformat=empty}
\usepackage{dcolumn}
\usepackage[]{natbib}
\bibliographystyle{apalike}

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\raggedright

\textbf{Title:} Tree height and leaf drought tolerance traits shape
growth responses across droughts in a temperate broadleaf forest

\textbf{Authors:} Ian R. McGregor\textsuperscript{1,2}, Ryan
Helcoski\textsuperscript{1}, Norbert Kunert\textsuperscript{1,3}, Alan
J. Tepley\textsuperscript{1,4}, Erika B.
Gonzalez-Akre\textsuperscript{1}, Valentine Herrmann\textsuperscript{1},
Joseph Zailaa\textsuperscript{1,5}, Atticus E.L.
Stovall\textsuperscript{1,6,7}, Norman A. Bourg\textsuperscript{1},
William J. McShea\textsuperscript{1}, Neil Pederson\textsuperscript{8},
Lawren Sack\textsuperscript{9,10}, Kristina J.
Anderson-Teixeira\textsuperscript{1,3}*

\textbf{Author Affiliations:}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Conservation Ecology Center; Smithsonian Conservation Biology
  Institute; National Zoological Park, Front Royal, VA 22630, USA
\item
  Center for Geospatial Analytics; North Carolina State University;
  Raleigh, NC 27607, USA
\item
  Center for Tropical Forest Science-Forest Global Earth Observatory;
  Smithsonian Tropical Research Institute; Panama, Republic of Panama
\item
  Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta,
  Canada
\item
  Biological Sciences Department; California State University; Los
  Angeles, CA 90032, USA
\item
  Department of Environmental Sciences, University of Virginia,
  Charlottesville, VA 22903, USA
\item
  NASA Goddard Space Flight Center; Greenbelt, MD 20771, USA
\item
  Harvard Forest, Petersham, MA 01366, USA
\item
  Department of Ecology and Evolutionary Biology; University of
  California, Los Angeles; Los Angeles, CA 90095, USA
\item
  Institute of the Environment and Sustainability; University of
  California, Los Angeles; Los Angeles, CA 90095, USA
\end{enumerate}

*corresponding author:
\href{mailto:teixeirak@si.edu}{\nolinkurl{teixeirak@si.edu}}; +1 540 635
6546

\begin{longtable}[]{@{}llll@{}}
\toprule
\begin{minipage}[b]{0.35\columnwidth}\raggedright
Text\strut
\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\raggedright
word count\strut
\end{minipage} & \begin{minipage}[b]{0.22\columnwidth}\raggedright
other\strut
\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\raggedright
n\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.35\columnwidth}\raggedright
Total word count (excluding summary, references and legends)\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
5,365\strut
\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
No.~of figures\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
4 (all colour)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.35\columnwidth}\raggedright
Summary\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
198\strut
\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
No.~of Tables\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
3\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.35\columnwidth}\raggedright
Introduction\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
1,034\strut
\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
No of Supporting Information files\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
\#\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.35\columnwidth}\raggedright
Materials and Methods\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
1,945\strut
\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.35\columnwidth}\raggedright
Results\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
697\strut
\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.35\columnwidth}\raggedright
Discussion\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
1467\strut
\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.35\columnwidth}\raggedright
Acknowledgements\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
125\strut
\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

\newpage

\hypertarget{summary}{%
\subsubsection{Summary}\label{summary}}

\begin{itemize}
\item
  As climate change is driving increased drought frequency and severity
  in many forested regions around the world, mechanistic understanding
  of the factors conferring drought resistance in trees is increasingly
  important. The dendrochronological record provides a window through
  which we can understand how tree size and species' traits shape tree
  growth responses during droughts.
\item
  We analyzed tree-ring records for twelve species that comprise 97\% of
  the woody productivity of the 25.6-ha ForestGEO plot in a broadleaf
  deciduous forest of northern Virginia (USA) to test hypotheses on how
  tree height, microenvironment characteristics, and species' traits
  shaped drought responses across the three strongest regional droughts
  over a 60-year period (1950 - 2009).
\item
  Individual-level drought resistance decreased with tree height, which
  was strongly correlated with exposure to higher evaporative demand and
  solar radiation. The potentially greater rooting volume of larger
  trees did not confer an advantage in sites with low topographic
  wetness index. Resistance was greater among species whose leaves
  experienced less shrinkage upon desiccation and lost turgor (wilted)
  at more negative water potentials.
\item
  We conclude that tree height and leaf drought tolerance traits
  influence growth responses during drought, as recorded in the
  tree-ring record spanning historical droughts. Thus, these factors can
  be useful for predicting future drought responses under climate
  change.
\end{itemize}

\emph{Key words}: annual growth; crown exposure; drought; Forest Global
Earth Observatory (ForestGEO); leaf drought tolerance traits; temperate
broadleaf deciduous forest; tree height; tree-ring

\newpage

\hypertarget{introduction}{%
\subsubsection{Introduction}\label{introduction}}

Forests play a critical global role in climate regulation
\citep{bonan_forests_2008}, yet there remains enormous uncertainty as to
how the forest-dominated terrestrial carbon sink will respond to climate
change \citep{friedlingstein_climatecarbon_2006}. An important aspect of
this uncertainty lies with physiological responses of trees to drought
\citep{kennedy_implementing_2019}. In many forested regions around the
world, the risk of severe drought is increasing
\citep{trenberth_global_2014, dai_climate_2018}, often despite
increasing precipitation
\citep{intergovernmental_panel_on_climate_change_climate_2015, cook_unprecedented_2015}.
Droughts, intensified by climate change, have been affecting forests
worldwide and are expected to continue as one of the most important
drivers of forest change in the future
\citep{allen_global_2010, allen_underestimation_2015}. Understanding
forest responses to drought requires elucidation of how tree size,
microenvironment, and species' traits jointly influence individual-level
drought resistance, and the extent to which their influence is
consistent across droughts. Because the resistance and resilience of
growth to drought is linked to trees' probability of surviving drought
\citep{desoto_low_2020, liu_hydraulic_2019}, understanding growth
responses can also help elucidate which trees are most vulnerable to
drought-induced mortality. However, it has proven difficult to resolve
the many factors affecting tree growth during drought with available
forest census data, which only rarely captures extreme drought, and with
tree-ring records, which capture multiple droughts but usually only
sample a subset of a forest community, typically focusing on a single
species or the largest individuals.

Many studies have shown that within and across species, large trees tend
to be more affected by drought. Greater growth reductions for larger
trees were first shown on a global scale by \citet{bennett_larger_2015},
and subsequent studies have reinforced this finding
\citep[\emph{e.g.},][]{hacket-pain_consistent_2016, gillerot_tree_2020}.
It has yet to be resolved which of several potential underlying
mechanisms most strongly shape these trends in drought response. First,
tree height itself may be a primary driver. Taller trees face the
biophysical challenge of lifting water greater distances against the
effects of gravity and friction
\citep{mcdowell_relationships_2011, mcdowell_darcys_2015, ryan_hydraulic_2006, couvreur_water_2018}.
Vertical gradients in stem and leaf traits--including smaller and
thicker leaves (higher leaf mass per area, LMA), greater resistance to
hydraulic dysfunction (\emph{i.e.}, more negative water potential at
50\% loss of hydraulic conductivity, more negative P50), and lower
hydraulic conductivity at greater heights
\citep{couvreur_water_2018, koike_leaf_2001, mcdowell_relationships_2011}--enable
trees to become tall \citep{couvreur_water_2018}. Greater stem
capacitance (\emph{i.e.}, water storage capacity) of larger trees may
also confer resistance to transient droughts
{[}\citet{phillips_reliance_2003}*; \citet{scholz_hydraulic_2011}{]}.
Indeed, tall trees require xylem of greater hydraulic efficiency, such
that xylem conduit diameters are wider in the basal portions of taller
trees, both within and across species
\citep{olson_plant_2018, liu_hydraulic_2019}, and throughout the
conductive systems of angiosperms
\citep{zach_vessel_2010, olson_universal_2014, olson_plant_2018}. Wider
xylem conduits plausibly make large trees more vulnerable to embolism
during drought \citep{olson_plant_2018}, and traits conducive to
efficient water transport may also lead to poor ability to recover from
or re-route water around embolisms \citep{roskilly_conflicting_2019}.

Larger trees may also have lower drought resistance because of
microenvironmental and ecological factors. Their crowns tend to occupy
more exposed canopy positions, which are associated with higher
evaporative demand \citep{kunert_revised_2017}. Subcanopy trees tend to
fare better specifically due to the benefits of a buffered environment
\citep{pretzsch_drought_2018}. Counteracting the liabilities associated
with tall height, large trees tend to have larger root systems
\citep{enquist_global_2002}, potentially mitigating some of the
biophysical challenges they face by allowing greater access to water.
Larger root systems--if they grant access to deeper water sources--would
be particularly advantageous in drier microenvironments (e.g., hilltops,
as compared to valleys and streambeds) during drought. Finally, tree
size-related responses to drought can be modified by species' traits and
their distribution across size classes
\citep{meakem_role_2018, liu_hydraulic_2019}. Understanding the
mechanisms driving the greater relative growth reductions of larger
trees during drought requires sorting out the interactive effects of
height and associated exposure, root water access, and species' traits.

Debates have also arisen regarding the traits influencing tree growth
responses to drought. Studies within temperate broadleaf forests have
observed ring-porous species showing higher drought tolerance than
diffuse-porous species
\citep{friedrichs_species-specific_2009, elliott_forest_2015, kannenberg_linking_2019},
but this distinction would not hold in the global context
\citep{wheeler_variations_2007, olson_xylem_2020} and does not resolve
differences among the many species within each category.
Commonly-measured traits including wood density and leaf mass per area
(\(LMA\)) have been linked to drought responses within some temperate
deciduous forests
\citep{abrams_adaptations_1990, guerfel_impacts_2009, hoffmann_hydraulic_2011, martinbenito_convergence_2015}
and across forests worldwide \citep{greenwood_tree_2017}. However, in
other cases these traits could not explain drought tolerance
\citep[e.g., in a tropical rainforest;][]{marechaux_leaf_2019}, or the
direction of response was not always consistent. For instance, higher
wood density has been associated with greater drought resistance at a
global scale \citep{greenwood_tree_2017}, but correlated negatively with
tree performance during drought in a broadleaf deciduous forest in the
southeastern United States \citep{hoffmann_hydraulic_2011}. Thus, the
perceived influence of these traits on drought resistance may actually
reflect indirect correlations with other traits that more directly drive
drought responses \citep{hoffmann_hydraulic_2011}.

In contrast, hydraulic traits have direct physiological linkages to tree
growth and mortality responses to drought. For instance, water
potentials at which percent the loss of conductivity surpasses a certain
threshold (e.g., P50 and P88, representing 50 and 88\% loss of
conductivity, respectively) and hydraulic safety margin (\emph{i.e.},
difference between typical minimum water potentials and P50 or P88)
correlate with drought performance across global forests
\citep{anderegg_meta-analysis_2016}. However, these are time-consuming
to measure and therefore infeasible for predicting or modeling drought
responses in highly diverse forests (\emph{e.g.}, in the tropics). More
easily-measurable leaf drought tolerance traits that have direct linkage
to plant hydraulic function can explain variation in plant distribution
and function \citep{medeiros_extensive_2019}. These include leaf area
shrinkage upon desiccation \citep[\(PLA_{dry}\);][]{scoffoni_leaf_2014}
and the leaf water potential at turgor loss point (\(\pi_{tlp}\)),
\emph{i.e.,} the water potential at which leaf wilting occurs
\citep{bartlett_correlations_2016, zhu_leaf_2018}. Both traits correlate
with hydraulic vulnerability and drought tolerance as part of unified
plant hydraulic systems
\citep{scoffoni_leaf_2014, bartlett_correlations_2016, zhu_leaf_2018, farrell_does_2017}.
The abilities of both \(PLA_{dry}\) and \(\pi_{tlp}\) to explain tree
drought resistance remains untested.
